The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-21 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Comments by Jonnie Pober

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength. However, a visibility compression technique using Delay/Delay-Rate filtering is used for PAPER, which makes it hard to identify low-level RFI. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of the Moon of terrestrial transmitters complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially restricted to the centre of the Lunar disk, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}. Another study shows that objects in space of about a meter in size, such as satellites and space debris, may also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

@@

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995} and/or \textsc{wsclean} \citep{offringa-wsclean-2014}, the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format. The output is then directly readable by the most common astronomical software packages.

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. These will be referred to as the EoR low and high bands, respectively. In these observations, the selected EoR field is tracked by the primary beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz. The GLEAM observations are made in drift-scan mode, with the 5 bands rotated through in sequence on a 2-minute cadence.

@@

The poly-phase filter bank that performs the first separation into sub-bands introduces a 1.28~MHz periodic spectral signature. As shown in Fig.~\ref{fig:bandpass}, the poly-phase filter band-pass shape is hard to remove completely, because of aliasing each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~MHz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data are lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV. The horizontal gray line represents the false-positives rate of the RFI detection.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
\caption{Visibility amplitude distributions with logarithmic axes. The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.43$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
\label{fig:lognlogs}
\end{center}
\end{figure*}

@@

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$. The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N\propto S^{-1.43}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively. Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.

@@

>>>>

%\defcitealias{lofar-radio-environment}{O2013a}
The interferometric arrays Low-Frequency Array (LOFAR; \citealt{lofar-2013}), Precision Array to Probe the Epoch of Reionization (PAPER; \citealt{parsons-paper-eorlimit-2014}) and the Giant Metrewave Radio Telescope (GMRT; \citealt{the-gmrt-swarup-chapter-2013}) observe in approximately the same frequency range as the MWA. For each of these instruments, projects are ongoing to detect redshifted 21-cm signals from the Epoch of Reionization. The overlap in frequency gives an opportunity to compare the observatory sites and hardware designs with respect to the RFI impact. So far, initial observations with the MWA have produced scientific results without RFI issues \citep{hurley-walker-mwacs-2014, mckinley-fornaxa-2014, hindson-cluster-emission-2014}, which of course is not surprising given its remote location. \citet{parsons-paper-eorlimit-2014} have shown that the radio environment for PAPER in the Karoo desert of South Africa is sufficiently clean to reach with long integration a $(41 \textrm{mK})^2$ upper limit for the EoR brightness temperature at one scale and redshift, two orders of magnitude (in $\textrm{mK}^2$) away from the expected EoR signal strength. \citet{lofar-radio-environment} show that for regular observations the radio environment of LOFAR does not pose issues, even though LOFAR is located in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors reach near-thermal noise sensitivity in the first EoR long-integration images with LOFAR. They argue that RFI is not a limitation for further increasing the sensitivity, and \citet{offringa-rfi-distributions} concludes that with sufficient precautions, such as good receiver design, accurate detection methods, and high time and frequency resolutions, residual RFI is weak and averages down in a similar way to Gaussian noise. However, none of the EoR projects have yet processed enough data to reach the sensitivities required for a detection of EoR signals, and low-level RFI could potentially prevent such a detection.

Recently, experiments to use the Moon as a calibrator for EoR experiments have shown that the reflection of terrestrial transmitters by the Moon complicates such an experiment \citep{mckinley-moon-2013}. However, reflections are spatially restricted to the centre of the Lunar disk, and the high resolution of LOFAR allows the separation of the reflected and intrinsic power from the Moon \citep{vedantham-2014-todo}. Another study shows that objects in space of about a meter in size, such as satellites and space debris, may also reflect terrestrial transmissions with enough strength to be observable by the MWA \citep{tingay-space-debris-2013}. While tracking space debris is a useful asset, such reflected RFI can be a problem for EoR experiments, especially all-sky experiments that try to measure the global EoR signal \citep{vedantham-2014-todo}.

@@

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is run by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is run in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For data processing with the Common Astronomy Software Applications (\textsc{casa}; \citealt{casa-2008}), \textsc{miriad} \citep{miriad-sault-1995}, \textsc{wsclean} \citep{offringa-wsclean-2014} and/or the Fast Holographic Deconvolution software (\textsc{fhd}, \citealt{fhd-sullivan-2012}), the \textsc{cotter} output is set to either the \textsc{casa} or \textsc{uv-fits} format. The output is then directly readable by the most common astronomical software packages.

@@

The MWA EoR project observes primarily in the 138.9--197.7~MHz range, covering the HI 21-cm line with redshift 6.1--9.2. Because the MWA has a 30.72~MHz instantaneous bandwidth, this range has been split up so that each night observes with a central frequency of either 154.2 or 182.4~MHz, with an overlap of 2.6 MHz. These will be referred to as the EoR low and high bands, respectively. For the EoR observations, the beam formers are changed every 30~minutes to track the selected field with the primary beam, such that the sensitivity towards the field is maximal. The GLEAM survey covers 72.3--230.8~MHz split into 5 bands of 30.72~MHz. The GLEAM observations are made in drift-scan mode, with the 5 bands rotated through in sequence on a 2-minute cadence.

@@

The poly-phase filter bank that performs the first separation into sub-bands introduces a 1.28~MHz periodic spectral signature. As shown in Fig.~\ref{fig:bandpass}, the poly-phase filter band-pass shape is hard to remove completely, because of aliasing each sub-band is affected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~MHz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data are lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. In particular, the detected fraction of RFI over frequency is about 0.5\% higher in the edge channels after flagging 80~kHz of the edges. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.
\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observation nights except GLEAM 2014-03-17. The latter has been left out because it is affected by DTV. The horizontal gray line represents the false-positives rate of the RFI detection. The RFI fractions are consistently higher than the false-positives rate because of transient broad-band RFI.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
\caption{Visibility amplitude distributions with logarithmic axes. The dashed line represent fits to the function $N=\beta S^\alpha$ over a reasonably constant part of the tail (selected by eye) of the RFI distribution, resulting in $\alpha=-1.37$ for GLEAM, $\alpha=-4.47$ for EoR high, and $\alpha=-1.33$ for EoR low.}
\label{fig:lognlogs}
\end{center}
\end{figure*}

@@

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N\propto S^{-4.47}$. The GLEAM and EoR low-band observations do show a small power-law component but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N\propto S^{-1.37}$ and $N\propto S^{-1.33}$ for the GLEAM and EoR-low curves, respectively. Due to the small region over which the power laws hold, as well as the possible high error in the fits because of the subjective data selection, it is hard to infer if a uniform distribution of transmitters is observed.

@@

Projects that try to detect very weak signals, such as the EoR projects, have to be careful that a (non)detection has not been affected by RFI leakage. One consideration is the storage of data: the symptoms of leaked RFI are harder to identify once visibilities have been averaged or compression techniques such as Delay/Delay-Rate filtering \citep{parsons-paper-eorlimit-2014} have been applied. Storing more information allows better verification of a detection and, if necessary, increases the chance of successfully performing further RFI excision.
